Composable Concurrency Models

### 2016-09-22 paper/paper.tex
Finished outline, sort of

The goal of this research is to discover and explain common properties of concurrency models (the actor model, software transactional memory, etc.) and demonstrate possible conflicts between them, ultimately showing how they can work together. The research is scoped to concurrency management---while some of the findings may apply to parallel programming, the source material primarily deals with language-level constructs.

@@

It seems reasonable to assume that the audience already has an intuitive understanding of different types of concurrent operations (disk reads, file downloads, etc.). However, concepts like atomicity and deadlock avoidance may have to be explained.

@@

@@

@@

@@

@@

\subsection{Coroutines}

@@

@@

\section{Composability challenges}

@@

@@

@@

@@

>>>>

The goal of this research is to discover and explain common properties of concurrency models and demonstrate possible conflicts between them, ultimately showing how they can work together. The research is scoped to concurrency management---while some of the findings may apply to parallel programming, the source material primarily deals with language-level constructs.

@@

It seems reasonable to assume that the audience already has an intuitive understanding of different types of concurrent operations (disk reads, file downloads, etc.). However, concurrent programming concepts (atomicity, deadlock, etc.) may have to be explained.

@@

Explanation of why concurrency is desirable (and usually necessary in practice). Example of an application that relies heavily on concurrency (web browser, for instance).

@@

Formal definitions from Ziv et.\ al~\cite{Ziv2015} are likely to be useful (serializability and visibility are treated particularly thoroughly). Marr and D'Hondt~\cite{Marr2012}, approaching the topic from a semantic enforcement angle, discuss atomicity in terms of managed mutation and execution. Other supplementary resources to fill out basic concurrency concepts would be helpful.

\subsection{Concurrency}

Brief introduction to threads, processes, and cores. Possible discussion of concurrency primitives (coroutines, fibers, green threads).

\subsection{Atomicity}

Explanation of individual effects/mutual exclusion.

\subsection{Serializability}

Explanation of execution order as a requirement for correctness.

\subsection{Visibility}

Explanation of thread opacity and communication.

@@

Swalens et al.~\cite{Swalens2014} provide some good explanations of the following, but they may be too closely coupled to Clojure. Other sources may be useful for fleshing out definitions or exploring other models.

@@

Definition of atomics and atomic operations (compare-and-swap, fetch-and-increment, locking).

@@

Description of concurrency models that use message passing for thread communication rather than shared memory (actors, agents).

@@

Explanation of futures (and deferreds and promises) as proxies for some unknown result.

@@

Definition of software transactional memory as a model that allows concurrent tasks to write to shared memory.

\section{Composability criteria}

Ziv et al.~\cite{Ziv2015} provide specific formal defininitions for composability correctness (``safety'' is described more specifically in terms of abortable and abort-free serializability and ``liveness'' is referred to as deadlock avoidance). Other sources seem to prefer the safety and liveness generalizations.

\subsection{Safety}

Definition of safety as a guarantee of partial correctness (``nothing bad will happen'').

\subsection{Liveness}

Definition of liveness as a guarantee of progress and eventual termination (``something good will eventually happen'').

@@

Sources tend to survey a particular space and evaluate solutions within it. This section may require more sources or a narrower scope.

\subsection{Language-specific assistance}

Languages can attempt to prevent common errors. (This maybe isn't as muh a solution as a workaround---Clojure, for instance, will handle some potentially unsafe combinations correctly.)

\subsection{Low-level abstractions}

Lower-level concurrency protocols can serve as the basis of implementation for some models, simplifying composition.

@@

Eloquent restatement of previous material.

@@

Mom, the Academy, etc.

@@

\nocite{*}
